Investigating the influence of genre preferences on movie/series rating patterns

### 2024-03-21 gen/paper/paper.Rmd

In this paper we investigate consumers’ movie and series preferences
using five of IMDb's datasets containing movie data. First, these
datasets are loaded, cleaned, inspected, merged and cleaned. Next, the
data is analyzed using a regression analysis. This regression analysis
regresses all movie and serie genre's against the average rating's over
a period of 5 years. Additionally, the variables title, movieType and
time are used as control variables to improve the interpret ability of
the estimates.

@@

Genres are used as the explanatory variable in the analysis. However,
because they are nested within the original data set they are first
un-nested and then pivoted wider. As a result data with 27 dummies was
constructed that could assessed separately in the analysis.

@@

Given that the measurement of genre preferences over time is a construct
that is viable to many unobservables, fixed effects were added to
partially capture these unobservant. More specifically the type of title
(movie vs. series) and title were added. To give an example, it could be
that for movies, people prefer something with a little more drama than
they would for a series or vice-versa.

A point of concern, however, is that these average ratings are not
measured over time but are a fixed sum up to and including the data of
analysis. This means that if series fall out of favor and their rating
drops this remains unobservant. Given that the continuation of a series
is determined by its success (a second season is not produced if the
first one was a flop) the following assumption can be made: if a genre
is not preferred throughout the continuance of a series, the rating is
low in the starting year and thus unlikely to be significant.
Additionally, since IMDb was not around in 1925, average ratings for
movies and series originating from that time may be strongly biased
because a nice selection of the population may be watching these movies
in the first place.

@@

Last to see how preferences change over time we build on the assumption
that if a movie or series is a success they gain many positive ratings
in the beginning and less so over time. As a result, there is a peak
during (the first season) release allowing us to only use the startYear
as a means to calculate behavior over time. Yet, the individual years
are unlikely to have a remarkable effect on the genre preference trends
This is because the effects are so nuanced that they are unlikely to be
detected. Therefore, to solve this a five-year window was created that
captures all genre preferences for that period

@@

While homoskedacity and independence appeared to be moderate at best.

@@

Regardless of these results the decision was made to continue with the
analysis. This is because the r package fixest allows for modification
in the mixed model analysis that helps overcome these limitations. For
example, the package allows for clustering of different groups of
residuals. While this does not entirely mitigate the assumption concerns
it does help with them.

@@

The results indicated that there are different genre preferences over
time. Out of all twenty-five genres, only eighteen appeared to
significantly influence the average ratings on IMBd. These genres were
animation, biography, film-noir, fantasy, documentary, crime, comedy,
game-show, history, horror, musical, shorts, sci-fi, romance,
reality-tv, news, and war.

The genre history fell in favor in around the year 1960 and continued to
significantly (p~1960-1964~\<.01) relate to the average rating until
this day (p~2020-2024~ \<.001). While the genre film-noir stoped to
significantly relate to the average ratings around the year 1950
(p~1950-1954~=.06 , p~1955-1959~ =.49). The genres crime (p~2000-2004~
\<.01,p~2005-2009~ \<.01), romance (p~1940-1944~ =.03, p~1945-1949~
\<.01), horror(p~1925-1929~ =.04), game-show(p~2000-2004~ \<.01),
reality-tv(p~1970-1974~ \<.01), sci-fi(p~1935-1939~ \<.01), and
shorts(p~1995-1999~ \<.01, p~2005-2009~ \<.01) had a single period
bursts were they significantly related to the average ratings. Last, the
genres animation (p\~1940-1944 \<.01, p~1950-1954~ \<.01, p~1975-1979~
\<.01 -up until 2024), biography (p~1935-1939~ \<.01, p~1985-1989~ \<.01
-Up until 2024), comedy(p~1930-1934~ =.02, p~1950-1954~ =.04,
p~1990-1994~ =.01), documentary (p~1930-1934~ \<.01, p\~1945-1949 \<.01,
p\~1980-1984 \<.01 - up to and including p\~2015-2019 \<.01), fantasy
(p~1930-1934~ \<.01, p~1935-1939~ \<.01, p~1945-1949~ \<.01,
p~1970-1974~ =.03), musical(p~1960-1964~ =.09, p~1985-1989~ =.03,
p~1990-1994~ =.02, p~2020-2024~\<.01), and war(p~1930-1934~ =.02,
p~1945-1949~ \<.01, p~1960-1964~ =.04 up to and including p~1970-1974~
\<.01, p~1985-1989~ \<.01 up to and including p~2005-2009~ \<.01)
significantly were related to the average rating in burst with unrelated
periods in between

@@

For genres like war and film-noir the correlation could be a sign of the
times, and strongly guided by biases in the rating population. For
example, when the genre war was indicated to be significantly related to
average ratings it was the time the Second World War and the Cold War
were in full motion. Therefore people who have an interest in this topic
may like to watch propaganda movies from that time. As a result, the
average ratings are higher because a nice of interested people is
watching. Similar to film-noir, people who are interested in this type
of movie or series are watching these and rating them. Therefore, there
may be a strong instance of viewer bias present.

On the contrary, the genres action, adult, adventure, mystery, sport,
thriller, and western are all insignificant and do not appear to be
correlated to the average ratings on IMBd. A reason for this may be that
these elements have become expected or core components of movies and
series and therefore do not stand out. As a result, they may not
correlate with the average rating as much as elements of a biography may
do. It is not every day a movie or series is released that is based on
true events.

@@

This research illustrates that genre preferences may be related to the
average IMBd ratings and vary over time. While further research is
necessary to fine-tune these claims the current research does provide
insight into how raters' watch patterns appear to alter. For example, it
shows that the common ingredients like romance and action may not stand
out as much as biography and history might. Additionally, the results
showed that The results of this analysis can be, at best, interpreted as
correlational given the vast amount of assumptions and potential hidden
dangers of sampling biases (e.g., film-noir and war genres). While this
has been partially mitigated by loosening some of the underlying
assumptions and adding fixed effects, the fact remains that these appear
as trend observations. Therefore, the results do not necessarily
conclude that some genres lead to higher average ratings. However, the
results do indicate that there may be preferences over time that may
lead to higher average ratings.

@@

Yet, the investigation into these genres is only a starting point. Much
more can be learned and investigated. Here are three ways in which the
current research into how genres relate to average IMBd rating may be
extended:

@@

Second, researchers may benefit by including data from multiple sources.
Currently, the analysis is limited to IMBd data but there are also other
sources such as rottentomato's for example. By combining these data sets
the interpretation of the results becomes less biased by the data
platform. For example, it may be that people of higher age groups
typically review IMBd, and younger groups adopt commentators. By
broadening the data sources the generalization of the results increases.

>>>>

In this paper we investigate consumers’ movie and series preferences
using four of IMDb's datasets containing movie data. First, these
datasets are loaded, inspected, merged and cleaned. Next, the
data is analyzed using a regression analysis. This regression analysis
regresses all movie and series genres against the average ratings over
a period of 5 years. Additionally, the variables title, movieType and
time are used as control variables to improve the interpretability of
the estimates.

@@

Genres are used as the explanatory variable in the analysis. However,
because they are nested within the original data set they are first
un-nested and then pivoted wider. As a result data with 27 dummies was
constructed that could be assessed separately in the analysis.

@@

Given that the measurement of genre preferences over time is a construct
that is viable to many unobservables, fixed effects were added to the 
regression analysis to partially capture these unobservables. More 
specifically the type of title (movie vs. series) and title were added. 
To give an example, it could be that for movies people prefer something 
with a little more drama than they would for a series or vice-versa.

A point of concern, however, is that these average ratings are not
measured over time but are a fixed sum up to and including the data of
analysis. This means that if series fall out of favor and their rating
drops this remains unobservant. Given that the continuation of a series
is determined by its success (a second season is not produced if the
first one was a flop) the following assumption can be made: if a genre
is not preferred throughout the continuance of a series, the rating is
low in the starting year and thus unlikely to be significant.
Additionally, since IMDb was not around in 1925, average ratings for
movies and series originating from that time may be strongly biased
because a niche selection of the population may be watching these movies
in the first place.

@@

Last, to see how preferences change over time we build on the assumption
that if a movie or series is a success they gain many positive ratings
in the beginning and less so over time. As a result, there is a peak
during (the first season) release allowing us to only use the startYear
as a means to calculate behavior over time. Yet, the individual years
are unlikely to have a remarkable effect on the genre preference trends.
This is because the effects are so nuanced that they are unlikely to be
detected. Therefore, a five-year window was created that captures all 
genre preferences for that period of time.

@@

Homoscedasticity and independence appeared to be moderate at best.

@@

Regardless of these results the decision was made to continue with the
analysis. This is because the R package `fixest` allows for modification
in the mixed model analysis that helps overcome these limitations. For
example, the package allows for clustering of different groups of
residuals. While this does not entirely mitigate the assumption concerns
it does help with them.

@@

The results indicated that there are different genre preferences over
time. Out of all twenty-five genres, only eighteen appeared to
significantly influence the average ratings on IMDb. These genres were
animation, biography, film-noir, fantasy, documentary, crime, comedy,
game-show, history, horror, musical, shorts, sci-fi, romance,
reality-tv, news, and war.

The genre history fell in favor in around the year 1960 and continued to
significantly (p~1960-1964~\<.01) relate to the average rating until
this day (p~2020-2024~ \<.001). While the genre film-noir stoped to
significantly relate to the average ratings around the year 1950
(p~1950-1954~=.06 , p~1955-1959~ =.49). The genres crime (p~2000-2004~
\<.01,p~2005-2009~ \<.01), romance (p~1940-1944~ =.03, p~1945-1949~
\<.01), horror (p~1925-1929~ =.04), game-show (p~2000-2004~ \<.01),
reality-tv (p~1970-1974~ \<.01), sci-fi (p~1935-1939~ \<.01), and
shorts (p~1995-1999~ \<.01, p~2005-2009~ \<.01) had a single period
bursts where they significantly related to the average ratings. Last, the
genres animation (p\~1940-1944 \<.01, p~1950-1954~ \<.01, p~1975-1979~
\<.01 -up until 2024), biography (p~1935-1939~ \<.01, p~1985-1989~ \<.01
-Up until 2024), comedy (p~1930-1934~ =.02, p~1950-1954~ =.04,
p~1990-1994~ =.01), documentary (p~1930-1934~ \<.01, p\~1945-1949 \<.01,
p\~1980-1984 \<.01 - up to and including p\~2015-2019 \<.01), fantasy
(p~1930-1934~ \<.01, p~1935-1939~ \<.01, p~1945-1949~ \<.01,
p~1970-1974~ =.03), musical (p~1960-1964~ =.09, p~1985-1989~ =.03,
p~1990-1994~ =.02, p~2020-2024~\<.01), and war (p~1930-1934~ =.02,
p~1945-1949~ \<.01, p~1960-1964~ =.04 up to and including p~1970-1974~
\<.01, p~1985-1989~ \<.01 up to and including p~2005-2009~ \<.01)
were significantly related to the average ratings in bursts with unrelated
periods in between.

@@

For genres like war and film-noir the correlation could be a sign of the
times, and can be strongly guided by biases in the rating population. For
example, when the genre war was indicated to be significantly related to
average ratings it was the time the Second World War and the Cold War
were in full motion. Therefore people who have an interest in this topic
may like to watch propaganda movies from that time. As a result, the
average ratings are higher because a niche of interested people is
watching. Similar to film-noir, people who are interested in this type
of movie or series are watching these and rating them. Therefore, there
may be a strong instance of viewer bias present.

On the contrary, the genres action, adult, adventure, mystery, sport,
thriller, and western are all insignificant and do not appear to be
correlated to the average ratings on IMDb. A reason for this may be that
these elements have become expected or core components of movies and
series and therefore do not stand out. As a result, they may not
correlate with the average rating as much as elements of a biography may
do. It is not every day a movie or series is released that is based on
true events.

@@

This research illustrates that genre preferences may be related to the
average IMDb ratings and vary over time. While further research is
necessary to fine-tune these claims the current research does provide
insight into how raters' watch patterns appear to alter. For example, it
shows that the common ingredients like romance and action may not stand
out as much as biography and history might. 
The results of this analysis can be, at best, interpreted as correlational 
given the vast amount of assumptions and potential hidden dangers of 
sampling biases (e.g., film-noir and war genres). While this has been 
partially mitigated by loosening some of the underlying assumptions and 
adding fixed effects, the fact remains that these appear as trend 
observations. Therefore, the results do not necessarily conclude that some 
genres lead to higher average ratings. However, the results do indicate that 
there may be preferences over time that may lead to higher average ratings.

@@

Yet, the investigation into these genres is only a starting point. Much
more can be learned and investigated. Here are three ways in which the
current research into how genres relate to average IMDb rating may be
extended:

@@

Second, researchers may benefit by including data from multiple sources.
Currently, the analysis is limited to IMDb data but there are also other
sources such as Rotten Tomatoes for example. By combining these data sets
the interpretation of the results becomes less biased by the data
platform. For example, it may be that people of higher age groups
typically review on IMDb, and younger groups adopt commentators. By
broadening the data sources the generalization of the results increases.
